---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: Assessing the risk of contamination from hazardous sites due to flooding in North Carolina low-socioeconomic communities  
subtitle: https://github.com/tristen0708/EJ_Project
author: Tristen Townsend
abstract: "It is well known that hazardous waste sites tend to be more frequently sited in low-income communities, particularly communities of color. Proximity to hazardous waste sites have potential to create health risks, especially if communities are in regions more likely to be prone to flooding or extreme precipitation. This analysis serves to understand whether risks might exist for low-income communities in North Carolina. The number and type of sites in various counties has been analyzed, as have which areas have been experiencing increased flooding and extreme precipitation events in the past decade. 92 words"
fontsize: 12pt
mainfont: Times New Roman
---

<Information in these brackets are used for annotating the RMarkdown file. They will not appear in the final version of the PDF document>

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

<Setup the global options for the R chunks in your document>

<Note: set up autoreferencing for figures and tables in your document>

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages

library(sf)
library(tidyverse)
library(dplyr)
library(viridis)
library(cluster)
library(stringr)
library(fuzzyjoin)
library(gridExtra)
library(kableExtra)

# Set your ggplot theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

```


# Research Question and Rationale

There is significant research that indicates hazardous waste sites, especially those listed on the National Priorities List as a Superfund site, are disproportionately located in communities of color or low socioeconomic status (Burwell-Naney et al., 2013; Kramar, Anderson, Hilfer, Branden, Gutrich, 2018). In North Carolina, recent hurricanes have resulted in serious flooding in many parts of the state, creating concern as to whether Superfund sites and other hazardous waste sites were breached and might pose health effects to local communities. Given that natural disasters such as hurricanes and increased flooding is expect in North Carolina, it is important to understand if risks related to hazardous waste sites and flooding are posted to minority communities, as they tend to be low-capacity and less resilient to disasters. This information could be used by environmental justice leaders to advocate for policy changes or the implementation of safeguards to be put in place.

During this analysis, I am interested in seeing if the number of hazardous waste sites is significantly greater in areas with a large percent of minority resident or a large amount of those in poverty. Additionally, for the areas with the highest site counts, I would like to see if they are at risk for flood events based on peak stage values. Poverty and race data by county will be used, as well as site data and peak stage data.

\newpage

# Dataset Information

I am using multiple datasets to answer my research question. I have downloaded geospatial data on hazardous waste site locations from North Carolina Department of Environmental Quality. I have downloaded data with information about poverty and race from the U.S. Census Bureau's Small Area Income and Poverty Estimates (SAIPE) Program and (U.S. Census Bureau, 2018) and from the NC Budget and Management LINC data retrievel tool. Furthermore, I've obtained peak stage data for a few counties in North Carolina over that past five years from the U.S. Geological Survey.



<Information on how the dataset for this analysis were collected, the data contained in the dataset, and any important pieces of information that are relevant to your analyses. This section should contain much of same information as the README file for the dataset but formatted in a way that is more narrative.>

```{r table}
Poverty_NC <- read.csv("./Data/Processed/NC_Poverty_processed.csv")
Race <- read.csv("./Data/Processed/LINC_RaceData_2010.csv")
Peak.Stage <- read.csv("./Data/Processed/FilteredPeaks.csv")  

```

\newpage

# Exploratory Data Analysis and Wrangling


```{r load and explore}
Poverty_NC <- read.csv("./Data/Processed/NC_Poverty_processed.csv")
Race <- read.csv("./Data/Processed/LINC_RaceData_2010.csv")
Peak.Stage <- read.csv("./Data/Processed/FilteredPeaks.csv") 

Peak.Stage_processed <- Peak.Stage %>% select("county", "latitude_dd", "longitude_dd", "vdatum", "peak_date", "peak_stage")

Race1 <- mutate(Race, WhitePerc = Race$White/Race$Total.Pop)
Race2 <- mutate(Race1, BlackPerc = Race$Black/Race$Total.Pop)
Race3 <- mutate(Race2, NativePerc = Race$Native/Race$Total.Pop)
Race4 <- mutate(Race3, AsianPerc = Race$Asian/Race$Total.Pop)
Race5 <- mutate(Race4, HispanicPerc = Race$Hispanic/Race$Total.Pop)
Race6 <- mutate(Race5, OtherPerc = Race$Other/Race$Total.Pop)
Race7 <- mutate(Race6, Haw_PacIsPerc = Race$Haw_PacIs/Race$Total.Pop)
Race8 <- mutate(Race7, TwoPerc = Race$Two/Race$Total.Pop)
Race_Processed <- mutate(Race8, MinorityPerc = (Race$Black + Race$Native + Race$Asian + Race$Hispanic + Race$Other + Race$Haw_PacIs + Race$Two)/Race$Total.Pop)

#Explore race and poverty data
shapiro.test(Race_Processed$MinorityPerc) #p-value = 0.0005985
shapiro.test(Poverty_NC$Poverty_Percent_allages) #p-value 0.01881 

#From the output, both p-values are < 0.05 implying that the distribution of the data are significantly different from normal distribution. In other words, we cannot assume the normality for either data sets.

#Explore discharge data
Peak.Stage_avg <- Peak.Stage_processed %>% 
  group_by(county) %>%
  summarise(yearly_average = mean(peak_stage))
  
shapiro.test(Peak.Stage_avg$yearly_average) #p-value = 0.288 normal!
#From the output, the p-value is < 0.05 implying that the distribution of the data are significantly different from normal distribution. In other words, we cannot assume the normality for either data sets.

```

```{r exploratory graphs}
ggplot(Race_Processed, aes(x = MinorityPerc)) +
  geom_freqpoly() +
  scale_x_continuous(limits = c(0, 1))
qqnorm(Race_Processed$MinorityPerc); qqline(Race_Processed$MinorityPerc)

ggplot(Poverty_NC, aes(x = Poverty_Percent_allages)) +
  geom_freqpoly()
qqnorm(Poverty_NC$Poverty_Percent_allages); qqline(Poverty_NC$Poverty_Percent_allages)
```

#Read in Counties shapefile into an sf dataframe, filtering for just NC counties
```{r load}
#Read in Counties shapefile into an sf dataframe, filtering for just NC counties

NC_Counties_shp <- st_read(dsn = "./Data/Spatial/NC_Counties.shp") #Geospatial data for NC Counties
Landfills_shp <- st_read(dsn = "./Data/Spatial/ActivePermittedLandiflls.shp") #Active Landfills
IH_shp <- st_read(dsn = "./Data/Spatial/IH_Sites.shp") #Hazardous substance spill and disposal sites 
FRB_shp <- st_read(dsn = "./Data/Spatial/FRB_Sites.shp") #Superfund
BF_shp <- st_read(dsn = "./Data/Spatial/BF_Sites.shp") #Brownfields
RUST_shp <- st_read(dsn = "./Data/Spatial/RUST.shp") #Underground Storage Tanks
HW_shp <- st_read(dsn = "./Data/Spatial/HW_Sites.shp") #Hazardous Waste Resource Conservation and Recovery Act
```


```{r wrangling}
#Reveal the CRS of the counties features so they can be graphed with NC Counties shapefile
st_crs(NC_Counties_shp)
st_crs(Landfills_shp)
st_crs(IH_shp)
st_crs(FRB_shp)
st_crs(BF_shp)
st_crs(RUST_shp)
st_crs(HW_shp)


#There is one row with an incorecct location; This site will be omitted in order to proceed with mapping data
Landfills_shp_mod <- subset(Landfills_shp, !LocationID == "P1252")

#Filter RUST dataset for only high risk UST sites since this inforation is available and due to the large amount of sites
levels(RUST_shp$ConfRisk)
highrisk_RUST <- RUST_shp %>%
  filter(ConfRisk == "H")

#Join poverty data to county geospatial data
county_poverty_join <- NC_Counties_shp %>% 
  left_join(y = Poverty_NC,by = c("CO_NAME" =  "Name"))


#Count the number of sites in each county; In some, the names of counties are not in all capital letters. In order to address future problems with joins, some were converted to all capital letters.
Sitecount_Landfill <- count(Landfills_shp_mod, Landfills_shp_mod$County)

levels(IH_shp$SITECOUNTY) <- toupper(levels(IH_shp$SITECOUNTY)) #all caps
Sitecount_IH <- count(IH_shp, IH_shp$SITECOUNTY) 

Sitecount_FRB <- count(FRB_shp, FRB_shp$SITE_COUNT) 

levels(BF_shp$BF_County) <- toupper(levels(BF_shp$BF_County))
Sitecount_BF <- count(BF_shp, BF_shp$BF_County)

levels(highrisk_RUST$County) <- toupper(levels(highrisk_RUST$County))
Sitecount_RUST <- count(highrisk_RUST, highrisk_RUST$County)

Sitecount_HW <- count(HW_shp, HW_shp$LOC_COUNTY)


#Join count dataframes with race info. Needed to make county names in Race_Processed data set capitalized so left join would work. In one case, Race_Processed data needed to add a column with abbreviations in order to join to the RUST data set.
Sitecount_Landfill_join <- left_join(Sitecount_Landfill, Race_Processed, by = c("Landfills_shp_mod$County" = "County"))

levels(Race_Processed$County) <- toupper(levels(Race_Processed$County))
Sitecount_IH_join <- left_join(Sitecount_IH, Race_Processed, by = c("IH_shp$SITECOUNTY" = "County"))

Sitecount_FRB_join <- left_join(Sitecount_FRB, Race_Processed, 
  by = c("FRB_shp$SITE_COUNT" = "County"))

Sitecount_BF_join <- left_join(Sitecount_BF, Race_Processed, by = c("BF_shp$BF_County" = "County"))

Race_Processed_abv <- transform(Race_Processed, ABV = str_sub(Race_Processed$County, 1, 5))

Sitecount_RUST_join <- left_join(Sitecount_RUST, Race_Processed_abv, by = c("highrisk_RUST$County" = "ABV"))

Sitecount_HW_join <- left_join(Sitecount_HW, Race_Processed, by = c("HW_shp$LOC_COUNTY" = "County"))

```

\newpage

# Analysis
<Include R chunks for 3+ statistical tests (display code and output) and 3+ final visualization graphs (display graphs only).>

```{r statistical tests}


```

```{r final visualizations}

#Creating maps from basic data with counties
ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = Landfills_shp_mod, alpha = 0.6) +
 labs(fill = "% in Poverty") +
  ggtitle("Landfill Sites in North Carolina")

ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = IH_shp, alpha = 0.3, color = "black") +
 labs(fill = "% in Poverty") +
  ggtitle("IH Sites in North Carolina")

ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = FRB_shp, alpha = 0.6, color = "black") +
 labs(fill = "% in Poverty") +
  ggtitle("Superfund Sites in North Carolina")

ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = BF_shp, alpha = 0.4) +
 labs(fill = "% in Poverty") +
  ggtitle("Brownfield Sites in North Carolina")

ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = highrisk_RUST, alpha = 0.3, color = "black") +
 labs(fill = "% in Poverty") +
  ggtitle("RUST Sites in North Carolina")

ggplot() + 
  geom_sf(data = county_poverty_join, aes(fill=county_poverty_join$Poverty_Percent_allages)) +
  scale_fill_viridis(direction = -1) +
  geom_sf(data = HW_shp, alpha = 0.3, color = "black") +
 labs(fill = "% in Poverty") +
  ggtitle("Hazardous Waste Sites in North Carolina")

#Graph count data with race percent data
Landfill_plot <- 
  ggplot(Sitecount_Landfill_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7, color="blue") +
  labs(x="Site Count", y="Racial Minority Percent")

IH_plot <- 
  ggplot(Sitecount_IH_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7, color="blue") +
  labs(x="Site Count", y="Racial Minority Percent")

FRB_plot <- 
  ggplot(Sitecount_FRB_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7) +
  labs(x="Site Count", y="Racial Minority Percent")
FRB_plot

BF_plot <- 
  ggplot(Sitecount_BF_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7, color="blue") +
  labs(x="Site Count", y="Racial Minority Percent")

RUST_plot <- 
  ggplot(Sitecount_RUST_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7, color="blue") +
  labs(x="Site Count", y="Racial Minority Percent")

HW_plot <- 
  ggplot(Sitecount_HW_join, aes(x=n , y=MinorityPerc * 100)) +
  scale_y_continuous(limits=c(0, 100)) +
  geom_point(alpha=0.7, color="blue") +
  labs(x="Site Count", y="Racial Minority Percent")

```

<Include text sections to accompany these R chunks to explain the reasoning behind your workflow, rationale for your approach, and the justification of meeting or failing to meet assumptions of tests.>


\newpage

# Summary and Conclusions
<Summarize your major findings from your analyses. What conclusions do you draw from your findings? Make sure to apply this to a broader application for the research question you have answered.>


\newpage

#References

U.S. Census Bureau. (2018). Small Area Income and Poverty Estimates (SAIPE) Program. Retrieved from https://www.census.gov/programs-surveys/saipe/about.html

NC Budget and Management. (2019). LINC. Retrieved from https://www.osbm.nc.gov/facts-figures/linc
